%
% This is the acknowledgement file (acknowl.tex)
%
%\Acknowledgement
%\begin{center}
%    \vspace{3em}
%    \LARGE{ACKNOWLEDGEMENTS}   
%\end{center}

I would like to thank my thesis supervisor, Dr.~Bennett Maruca, for his constant support and
understanding during the whole duration of my graduate education. He has been a source of
inspiration and encouragement.

I would like to thank my colleagues in the department with whom I have had innumerable discussions
on all topics, academic and otherwise. A lot of ideas presented in this thesis originated during one
of those discussions. Coffee-shop meetings with Dr.~Bill Matthaeus was THE place for such
discussion. I am especially indebted to Tulasi Parashar and Rohit Chhiber, who were always very
patient while explaining nuances of plasma physics to me. Alexandros Chasapis was always encouraging
and informative as we discussed the subtleties of multi-spacecraft techniques. Discussion with
Riddhi, Mark, Debanjan, and Manuel were always animated, fun, educational and diverse. I am
specially indebted to Bill and Riddhi who always took time to explain finer points of turbulence. I
am so glad to have been part of such a dynamic and diverse group.

I feel lucky to have had the chance to work with Dr.~Peter Gary who was always excited about the
kind of work we were doing. His comments on our results were always insightful and encouraging and
his enthusiasm for discussion was contagious. He will be fondly remembered and dearly missed!

I am very thankful to the MMS and PSP teams for giving me the opportunity and most importantly data
to work with.

Life in Newark was filled with fun and I depart here with lots and lots of memories. Thanks to all
my friends who contributed to the joy of living there. Siddhi, Sucharita, Mehul, Abdul, Sameer,
Mark, Khushboo, Aditya, Abhishek, Prasanna, Anitha, Vrathasha; thank you all for making my stay so
much enjoyable.

Special thanks to Krishna who has been supportive and understanding throughout thesis writing
process and helped in typing a not so insignificant part of this thesis.

I would also like to thank the staff of the Department of Physics and Astronomy, especially Maura,
Elle, Angie and Linda, who have been extremely helpful over the years and very proactive in solving
my non-academic issues. Sansern Somboonsong and Zelphia Johnson were always very responsive to my
queries regarding the computer and network.

I would like to thank my family, who despite their own biases against science, have largely been
supportive my own endeavors in the field.

And lastly I would like to thank my dear friend Abhishek Maniyar. I decided to leave my job, and a
good job it was, because he insisted that I should and pursue my dream of a PhD in Physics. He
guided me through the whole process and I am extremely thankful to him for making me do that. Thank
you Maniyar for being the force behind this!

This research made use of PlasmaPy, a community-developed open source Python package for plasma
science \citep{misc:Plasmapy2020}. I extensively used the SAO/NASA
\href{https://ui.adsabs.harvard.edu/}{Astrophysics Data System (ADS)} in the preparation of this
dissertation.